---
title: "Final Project STAT442"
author: "Chaozhong Wang"
date: "2025-04-13"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Option 1

```{r}
library(tidyverse)
library(lubridate)
library(ggrepel)  # For better label placement

# Prepare data with correct quarterly dates
trade_data <- read_csv("dataset_2025-04-13T02_01_49.813546073Z_DEFAULT_INTEGRATION_IMF.STA_IMTS_1.0.0.csv") %>%
  separate(TIME_PERIOD, into = c("Year", "Quarter"), sep = "-Q", remove = FALSE) %>%
  mutate(
    Year = as.integer(Year),
    Quarter = as.integer(Quarter),
    Date = ymd(paste(Year, Quarter * 3 - 2, "01"))  # First day of quarter
  )

# Identify extreme countries (top/bottom 3 by average absolute trade balance)
extreme_countries <- c("United States",
                       "China, People's Republic of",
                       "Germany",
                       "India",
                       "United Kingdom",
                       "Vietnam"
  
)

# Create labeled time series plot
ggplot(trade_data %>% filter(COUNTRY %in% extreme_countries), 
       aes(x = Date, y = OBS_VALUE/1e6, color = COUNTRY)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
    scale_y_continuous(labels = scales::comma) +
  labs(title = "Trade Balance with Canada by Country (2021-2024)",
       x = "Year",
       y = "Trade Balance (Millions USD)",
       color = "Country")
```

The line plot gives us an overview of the trade balance data over the past 4 years.

Canada's trade relationships show striking imbalances, with the United States dominating deficits and China leading surpluses. 

The data reveals two distinct tiers:

Extreme volatility: US-China trade swings exceed $10B quarterly;

Stable partners: Vietnam/Germany maintain consistent surpluses.




\newpage

## Option 2

```{r geographical}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(maps)        # For world map data
library(countrycode) # For country name standardization
library(sf)          # For spatial data handling
library(rnaturalearth) # For high-resolution world maps
library(viridis)     # For color scales

data = read_csv("dataset_2025-04-13T02_01_49.813546073Z_DEFAULT_INTEGRATION_IMF.STA_IMTS_1.0.0.csv")
# Standardize country names to match map data
data <- data %>%
   mutate(COUNTRY = case_when(
     COUNTRY == "China, People's Republic of" ~ "China",
     COUNTRY == "Hong Kong Special Administrative Region, People's Republic of" ~ "Hong Kong",
     COUNTRY == "Netherlands, The" ~ "Netherlands",
     COUNTRY == "Egypt, Arab Republic of" ~ "Egypt",
     TRUE ~ COUNTRY
   ))

# #add continent information
# data <- data %>%
#   mutate(continent = countrycode(COUNTRY, "country.name", "continent",
#                               custom_match = c("Hong Kong" = "Asia")))

# Get world map data
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(name != "Antarctica")

#repare trade data for mapping (using average trade balance for each country)
trade_summary <- data %>%
  group_by(COUNTRY) %>%
   summarize(avg_trade = mean(OBS_VALUE/1e6, na.rm = TRUE)) %>% ungroup()

# Join trade data with map data
world_trade <- world %>%
  left_join(trade_summary, by = c("name_long" = "COUNTRY"))


# Choropleth Map of Average Trade Balance
ggplot(world_trade) +
  geom_sf(aes(fill = avg_trade), color = "white", size = 0.2) +
  scale_fill_gradient2(low = "yellow", mid = "green", high = "blue",
                      midpoint = 0, 
                      labels = scales::comma,
                      name = "Avg Trade Balance\n(Millions USD)") +
  labs(title = "Average Trade Balance with Canada for Selected Countries (2021-2024)",
       subtitle = "Green = Trade Surplus with Canada | Yellow = Trade Deficit") +
  theme_minimal() +
  theme(legend.position = "right",
        axis.text = element_blank(),
        panel.grid = element_blank())


```

China has a trade surplus with Canada an average over 2021 - 2024.

The United States has a trade deficit with Canada an average over 2021 - 2024.

Most countries have a trade balance around 0 with Canada.




\newpage

## Option 3

```{r categorical}
#by continent

library(tidyverse)
library(ggplot2)
library(countrycode)
library(viridis)
library(ggrepel) # For better label placement


data <- data %>%
    separate(TIME_PERIOD, into = c("Year", "Quarter"), sep = "-Q", remove = FALSE) %>%
  mutate(
    Year = as.integer(Year),
    Quarter = as.integer(Quarter),
    Date = ymd(paste(Year, Quarter * 3 - 2, "01"))  # First day of quarter
  ) %>%
  mutate(continent = countrycode(COUNTRY, "country.name", "continent",
                               custom_match = c(
                                 "Hong Kong" = "Asia",
                                 "Kosovo" = "Europe",
                                 "Taiwan" = "Asia"))) %>%
  # Handle any remaining NA continents
  mutate(continent = case_when(
    COUNTRY == "European Union" ~ "Europe",
    TRUE ~ continent
  )) %>%
  filter(!is.na(continent)) # Remove any countries without continent classification
continent_palette <- c("#440154FF", "#433E85FF", "#2E6F8EFF", "#1F9A8AFF", "#4DC36BFF", "#BBDF27FF")
# Stacked Area Chart of Continental Aggregates
continent_totals <- data %>%
  group_by(Date, continent) %>%
  summarize(total_trade = sum(OBS_VALUE, na.rm = TRUE)/1000) %>%
  mutate(continent = fct_reorder(continent, -total_trade, .fun = sum))

ggplot(continent_totals, aes(x = Date, y = total_trade, fill = continent)) +
  geom_area(alpha = 0.8) +
  scale_fill_manual(values = continent_palette) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Aggregate Trade Balance with Canada by Continent",
       subtitle = "Stacked area shows combined trade balance for all countries in each continent",
       x = "Year",
       y = "Total Trade Balance (Millions USD)",
       fill = "Continent") +
  theme_minimal() +
  theme(legend.position = "bottom")
```


Canada consistently runs trade deficits with its continental neighbors in the Americas, particularly with the United States and Mexico, as shown by negative trade balances. 

Meanwhile, trade surpluses with Asian nations continue to grow, led by China and Vietnam, reflecting Asia's role as a manufacturing powerhouse and key exporter to the Canadian market.

\newpage

## Option 5

```{r table}
library(tidyverse)
library(knitr)
library(formattable)

# Read and prepare data with volatility calculations
volatility_data <- read_csv("CAtrading(countryperrow).csv") %>%
  pivot_longer(
    cols = `2021-Q1`:`2024-Q4`,
    names_to = "Quarter",
    values_to = "Trade_Balance"
  ) %>%
  group_by(COUNTRY, CONTINENT) %>%
  summarise(
    Avg_Balance = mean(Trade_Balance),
    Volatility = sd(Trade_Balance),
    Min_Balance = min(Trade_Balance),
    Max_Balance = max(Trade_Balance),
    .groups = "drop"
  ) %>%
  mutate(
    Volatility_Level = case_when(
      Volatility > 1500 ~ "Extreme",
      Volatility > 500 ~ "High",
      Volatility > 100 ~ "Moderate",
      TRUE ~ "Low"
    ),
    Balance_Type = ifelse(Avg_Balance >= 0, "Surplus", "Deficit")
  ) %>%
  arrange(desc(Volatility)) 
colnames(volatility_data) <- c(
  "COUNTRY",
  "CONTINENT",
  "Avg Balance",
  "Volatility",
  "Min Balance",
  "Max Balance",
  "Volatility Level",
  "Balance Type"
)
# Create conditional formatting table
formattable(volatility_data, list(
  `COUNTRY` = formatter("span", style = ~ style(color = "grey", font.weight = "bold")),
  `Volatility` = color_tile("white", "red"),
  `Avg Balance` = formatter("span",
    style = x ~ style(
      color = ifelse(x >= 0, "green", "red"),
      font.weight = ifelse(abs(x) > 1000, "bold", "normal")
    ),
    x ~ accounting(x, digits = 0)
  ),
  `Volatility Level` = formatter("span",
    style = x ~ style(
      display = "block",
      padding = "0 4px",
      "border-radius" = "4px",
      "background-color" = case_when(
        x == "Extreme" ~ "#FF0000",
        x == "High" ~ "#FF6B6B",
        x == "Moderate" ~ "#FFD166",
        TRUE ~ "#CCE5FF"
      )
    )
  ),
  `Balance Type` = formatter("span",
    style = x ~ style(
      display = "block",
      padding = "0 4px",
      "border-radius" = "4px",
      "background-color" = ifelse(x == "Surplus", "#C7F5C7", "#FFC7C7")
    ))),
  caption = "Trade Balance Volatility with Canada (2021-2024), unit in Million")
```

Countries are sorted by volatility, the amount of change.
U.S. has the most volatility, followed by China.






